---
title: "Activity_Tracking"
author: "Apurv Kaushal"
date: "16 December 2018"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment Description

This is the reprot for the Week 2 assignment for the Reproducible Data Research Course under the John Hopkins Data Science Specialisation at Coursera.

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

## Loading & Pre Processing Data
```{r process}
activities<-read.csv("C:\\Users\\Apurv\\Desktop\\Data_Science\\activity.csv")
str(activities)
library(dplyr)
numberNA<-sum(is.na(activities$steps))
frac_na<-numberNA/nrow(activities)
activity<-filter(activities,!is.na(activities$steps))
```
In the above code, data was loaded & preprocessed to understand the statistics of the various data features and isolate dataset with no missing values.

## Mean Number of Steps Per Day
```{r mean_no}
activity_aggr<-aggregate(activity$steps,by=list(dates=activity$date),FUN=sum)
hist(activity_aggr$x,xlab="days",ylab="Total steps",col="blue",main="Total steps per day")
summary(activity_aggr$x)
```

In the above code, the total number of steps per day was calculated & plotted as a histogram. The mean & median of the dataset with no missing values is 10766 & 10765 respectively.

## Daily Activity Pattern
```{r daily pattern}
activity_aggr2<-aggregate(activity$steps,by=list(interv=activity$interval),FUN=mean)
plot(activity_aggr2$x~activity_aggr2$interv,type="l",xlab="5 minute interval",ylab="Steps avg across days")
```

In the above code, the line plot is between the 5 minute intervals of a day & the steps during that interval, averaged across all days. The time interval 8:35-8:40 has the maximum number of steps, probably due to the active morning routines of people.

## Imputing Missing Values
```{r impute}
print("Total NA values ")
numberNA
activity_miss<-filter(activities,is.na(activities$steps))
for (i in 1:nrow(activity_miss))
  activity_miss[i,]$steps<-activity_aggr2[activity_aggr2$interv==activity_miss[i,]$interval,]$x
activity_imputed<-rbind(activity,activity_miss)
activity_imputed<-activity_imputed[order(activity_imputed$date),]
activity_aggr3<-aggregate(activity_imputed$steps,by=list(dates=activity_imputed$date),FUN=sum)
hist(activity_aggr3$x,xlab="days",ylab="Total steps",col="green",main="Total steps per day")
summary(activity_aggr3$x)
```
The strategy used for imputing is replacing the NA values with the number of steps averaged over days, in that time interval of the day. The Total number of missing values in the initial dataset is 2304. As expected, the total number of steps per day increased as the imputed values now replaced the missing values in each day. The mean & median of this imputed dataset is 10766 and 10766 respectively. The mean did not change because the imputed values were derived from the mean values of the dataset with non missing values itself. The median increased by 1.

## Difference between weekday & weekend
```{r weekdaynend}
activity_imputed$daytype<-weekdays(as.Date(activity_imputed$date))
weekday_act<-filter(activity_imputed,activity_imputed$daytype!="Saturday" & activity_imputed$daytype!="Sunday")
weekd_aggr<-aggregate(weekday_act$steps,by=list(interv=weekday_act$interval),FUN=mean)
weekend_act<-filter(activity_imputed,activity_imputed$daytype=="Saturday" | activity_imputed$daytype=="Sunday")
weeke_aggr<-aggregate(weekend_act$steps,by=list(interv=weekend_act$interval),FUN=mean)
par(mfrow=c(2,1))
plot(weekd_aggr$x~weekd_aggr$interv,type="l",xlab="weekday_interval",ylab="Steps avg across weekdays")
plot(weeke_aggr$x~weeke_aggr$interv,type="l",xlab="weekend_interval",ylab="Steps avg across weekend")
```

The plots of weekends indicate a much more active routine in comparison to the weekdays for hours post 10 AM in the morning. This is an indication of shopping or going out behaviour, often opted out during the weekends in comparison to the popular desk jobs at work. The time before 10 AM, the number of steps during the weekdays seem to be a little higher atleast around the 8 hour mark. On weekdays, people generally tend to wake up a little later as there is no work.